Rising Trends in Shelter Resident Deaths in Toronto Over the Past 17 Years

### 2024-09-25 paper/paper.qmd
Bugs fixed, Paper Done, Formatted
1. Bugs in test file fixed.
2. Paper finished.
3. Graphs, tables modified.
4. and other minor problem fixed

---
title: "Rising Trends in Shelter Resident Deaths in Toronto Over the Past 17 Years"
author: 
  - Haowei Fan
date: today
date-format: long
abstract: "Shelter resident deaths are an important indicator of social stability, reflecting systemic issues such as welfare systems, housing crises, and healthcare access. This project analyzed monthly shelter resident deaths in the Toronto region from March 2007 to August 2024, focusing on trends over time and seasonal variations. The study found that deaths have been increasing, with a significant rise between 2020 and 2023. While the number of deaths remained similar in spring, summer, and fall, winter months saw a significantly higher number of fatalities. These findings provide valuable data to support targeted interventions by governments and welfare organizations to address changing needs throughout the year."
format: pdf
number-sections: true
bibliography: references.bib

@@

```{r}
#| include: false
#| warning: false
#| message: false
#| echo: false
#### Workspace setup ####
library(tidyverse)
library(knitr)
#### Read data ####
analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
```

@@

While many sociological studies have explored the growing number of shelter residents. For example, Maureen Crane and Louise Joly have studied the growing number of shelter residents and the issue of aging in Western countries (@Crane_Joly_2014). Similarly, in their article, John M. Quigley, Steven Raphael, and Eugene Smolensky pointed out that the number of shelter residents in the United States is increasing, along with the underlying causes contributing to this rise (@AmericanHomeless). Fewer have examined shelter resident mortality rates in detail. Given the challenges of accurately counting the transient and often untracked shelter population, studying actual death counts offers a more reliable and feasible method of understanding the impact of systemic issues. Mortality data, being both concrete and verifiable, can better inform policy decisions than estimates of shelter population numbers, which are often difficult to obtain and track (@LackAppraoch).

This paper focuses on analyzing the number of deaths among shelter residents in Toronto from March 2007 to August 2024. The aim is to identify both long-term trends and seasonal variations in shelter resident deaths. Using simple linear regression, this study estimates the rate at which the number of deaths has increased over time. This project also examines anomalies in the data by comparing yearly death numbers, and assesses seasonal differences by analyzing average death rates across the four seasons. This analysis provides valuable insights into the hardships faced by shelter residents, especially during specific times of the year.

This study finds three key trends: 1) the number of deaths among shelter residents in Toronto has been rising over time; 2) there was a noticeable increase in deaths between 2020 and 2023; and 3) while deaths during spring, summer, and fall remained relatively consistent, winter saw significantly higher mortality rates. Understanding these trends is critical because shelter resident deaths serve as a key indicator of broader social problems, such as gaps in welfare systems, housing shortages, and insufficient healthcare services. These findings provide essential data for governments and welfare organizations to develop more targeted interventions based on the needs of shelter residents throughout the year.

@@

The dataset I am using is titled "Deaths of Shelter Residents", sourced from OpenToronto (@Opendatatoronto). This dataset records the number of deaths among shelter residents in the Toronto area, starting from January 2007 on a monthly basis. It also includes the gender of the deceased, categorized into three sections: male, female, and other gender, which includes Transgender, Non-binary, and Two-Spirit (@Opendatatoronto). The dataset is continuously updated. @tbl-column contains descriptions of these variables.

@@

# Draw the table.
kable(column_description, col.names = c("Column", "Description"))
```

In addition to this dataset, OpenToronto also offers a similar dataset called "Deaths of People Experiencing Homelessness", which includes four variables: id, year, month, and number of deaths (@Opendatatoronto). However, I chose not to use this dataset due to its limited data coverage, spanning only from May 2007 to August 2023. In comparison, the "Deaths of Shelter Residents" dataset provides more extensive data, allowing for greater accuracy in the results.

@@

The data cleaning process involved several key steps, each designed to improve the dataset's suitability for analysis. First, the year and month variables were combined, with the month converted to numerical format to facilitate time series analysis. A new variable was then created to classify each month into one of the four seasons: March, April, and May as spring; June, July, and August as summer; September, October, and November as autumn; and December, January, and February as winter, enabling seasonal analysis. To streamline the dataset, unnecessary variables, such as ID and the three gender categories, were removed. Additionally, to ensure that the seasonal analysis of average shelter resident deaths covered equal time intervals for each season, certain samples from the dataset's beginning and end were removed, resulting in a total number of samples divisible by four. Finally, after reviewing the dataset, it was confirmed that there were no missing values, so no further treatment was necessary. Upon completion, the cleaned dataset was reduced to three variables: Time, Total Deaths, and Season, as outlined in @tbl-sample.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-sample
#| tbl-cap: "Sample of the cleaned data"

# Display few of samples.
kable(head(analysis_data))
```

## Statistical summary of the cleaned data
@tbl-sumarry provides some basic statistics about the dataset. The project spans a total of 204 months. During this period, the highest number of deaths recorded in a single month was 19, while the lowest was 0. The average number of deaths per month is 3.848. Additionally, we can infer that the data is somewhat skewed, as the highest number of deaths in a single month is 19, while the average and median number of deaths are 3 and 3.83, respectively.
```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-sumarry
#| tbl-cap: "Basic statistics of the cleaned data"

# Showing the summary of cleaned data.
kable(summary(analysis_data[, c("year_month", "total_decedents")]))
```

@fig-deathovertime depicts the monthly variation in the number of deaths over time. The overall trend reveals a significant increase in the total number of shelter resident deaths in Toronto, particularly after 2015. Prior to 2015, the number of deaths remained relatively stable, rarely exceeding 5 per month. However, from 2015 onwards, there is not only a noticeable upward trend in deaths but also increased volatility. Spikes in the data become more frequent, with several months recording over 10 deaths, and some months reaching as high as 15 or more.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-deathovertime
#| fig-cap: "Total Decedents Over time"

# Line plot for total_decedents over time
ggplot(analysis_data, aes(x = as.Date(paste(year_month, "-01", sep="")), y = total_decedents)) +
  geom_line() +
  labs(title = "Total Decedents Over Time", x = "Date", y = "Total Decedents") +
  theme_minimal()
```

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| include: false
# Convert year_month to a date format, assuming year_month is a character string in the format YYYY-MM
analysis_data$year_month_date <- as.Date(paste0(analysis_data$year_month, "-01"))

# Set the start date to 2007-01-01
start_date <- as.Date("2007-01-01")

# Calculate the time difference (in weeks) between each date and the start date
analysis_data$days_since_start <- as.numeric(difftime(analysis_data$year_month_date, start_date, units = "weeks"))

# Perform a regression analysis using the calculated time difference
m <- lm(total_decedents ~ days_since_start, data = analysis_data)
```
I used the total number of shelter resident deaths per month as the target variable to be predicted, with time and season as predictors, and performed a linear regression analysis. Since time and season are evidently highly correlated, I decided to retain only time as the final predictor for the model. After completing the simple linear regression, I obtained a positive slope, represented by a straight line (@fig-slr), indicating that the number of shelter resident deaths in Toronto has increased over time. A comparison with the actual data revealed that the real values closely align with the predictions from the SLR model, demonstrating the reliability of my model.
```{r, results='hide'}
#| warning: false
#| message: false
#| echo: false
#| label: fig-slr
#| fig-cap: "Simple linear regression"

# Scatter plot and fitted line
ggplot(analysis_data, aes(x = ym(year_month), y = total_decedents)) +
  geom_point() +
  geom_line(aes(y = fitted(m)), col = "blue", size = 1)
  labs(title = "Total Decedents vs Year/Month",
       x = "Time",
       y = "Total Decedents")
```

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-yeardeath
#| fig-cap: "Total Decedents per Year"

@@

# Create a bar chart with percentage labels
ggplot(filtered_data, aes(x = year, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_text(aes(label = paste0(round(total_decedents_pct, 1), "%")), 
            vjust = -0.5, size = 3) +  # Add percentage labels
  labs(title = "Total Decedents per Year", x = "Year", y = "Total Decedents") +
  theme_minimal()

```

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-seasondeath
#| fig-cap: "Average Decedents by Season"

@@

# Create a bar chart with percentage labels
ggplot(seasonal_data, aes(x = season, y = mean_decedents, fill = season)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(total_decedents_pct, 1), "%")), 
            vjust = -0.5, size = 3) +  # Add percentage labels
  labs(title = "Average Decedents by Season", 
       x = "Season", 
       y = "Average Decedents") +
  theme_minimal()
```

>>>>

---
title: "Rising Trends in Shelter Resident Deaths in Toronto Over the Past 17 Years"
author: 
  - Haowei Fan
date: today
thanks: "Code and data are available at: https://github.com/HaoweiFan0912/Deaths-of-Shelter-Residents-in-Toronto.git"
date-format: long
abstract: "Shelter resident deaths are an important indicator of social stability, reflecting systemic issues such as welfare systems, housing crises, and healthcare access. This project analyzed monthly shelter resident deaths in the Toronto region from March 2007 to August 2024, focusing on trends over time and seasonal variations. The study found that deaths have been increasing, with a significant rise between 2020 and 2023. While the number of deaths remained similar in spring, summer, and fall, winter months saw a significantly higher number of fatalities. These findings provide valuable data to support targeted interventions by governments and welfare organizations to address changing needs throughout the year."
format: pdf
number-sections: true
bibliography: references.bib

@@

```{r}
#| include: false
#| warning: false
#| message: false
#| echo: false
#### Workspace setup ####
library(tidyverse)
library(knitr)
library(dplyr)
set.seed(912)
#### Read data ####
analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
```

@@

While many sociological studies have explored the growing number of shelter residents. For example, Maureen Crane and Louise Joly have studied the growing number of shelter residents and the issue of aging in Western countries (@Crane_Joly_2014). Similarly, in their article, John M. Quigley, Steven Raphael, and Eugene Smolensky pointed out that the number of shelter residents in the United States is increasing, along with the underlying causes contributing to this rise (@AmericanHomeless). Fewer have examined shelter resident mortality rates in detail. Given the challenges of accurately tracking the transient shelter population, studying actual death counts—concrete and verifiable data—offers a more reliable and feasible method for understanding changes in the needs of shelter residents, providing better insights for policy decisions than difficult-to-obtain population estimates (@LackAppraoch).

This paper analyzes the number of deaths among shelter residents in Toronto from March 2007 to August 2024, aiming to identify long-term trends and seasonal variations. Using simple linear regression, the study estimates the rate of increase in deaths over time, while also examining anomalies by comparing yearly death numbers and assessing seasonal differences through average death rates across the four seasons. The findings reveal three key trends: 1) a steady rise in the number of shelter resident deaths over time; 2) a noticeable spike in deaths between 2020 and 2023; and 3) significantly higher mortality rates during winter compared to the relatively consistent rates in spring, summer, and fall. These insights provide critical data for governments and welfare organizations to develop more targeted interventions based on the needs of shelter residents throughout times.

@@

The dataset used in this study is titled "Deaths of Shelter Residents" and is sourced from OpenDataToronto (@Opendatatoronto). It was published by Toronto Public Health (TPH) and records the number of deaths among shelter residents in the Toronto area on a monthly basis, starting from January 2007. Additionally, the dataset includes the gender distribution of the deceased each month, with gender categorized into three groups: male, female, and other gender, which encompasses Transgender, Non-binary, and Two-Spirit individuals (@Opendatatoronto). The dataset is continuously updated. A detailed description of the variables can be found in @tbl-column.

@@

# Draw the table.
kable(column_description, col.names = c("Variables", "Descriptions"))
```

In addition to this dataset, OpenDataToronto also provides a similar dataset titled "Deaths of People Experiencing Homelessness," which records monthly deaths of homeless individuals from January 2017 to March 2023 (@Opendatatoronto). However, I chose not to use this dataset due to its limited coverage, spanning only from January 2017 to December 2023. In comparison, the "Deaths of Shelter Residents" dataset offers a longer time frame, allowing for more comprehensive analysis and greater accuracy in the results (@Opendatatoronto).

@@

The data cleaning process involved several key steps, each designed to improve the dataset's suitability for analysis. First, the year and month variables were combined, with the month converted to numerical format to facilitate time series analysis. A new variable was then created to classify each month into one of the four seasons: March, April, and May as spring; June, July, and August as summer; September, October, and November as autumn; and December, January, and February as winter, enabling seasonal analysis. To streamline the dataset, unnecessary variables, such as ID and the three gender categories, were removed. Additionally, to ensure that the seasonal analysis of average shelter resident deaths covered equal time intervals for each season, certain samples from the dataset's beginning and end were removed, resulting in a total number of samples divisible by four. Finally, after reviewing the dataset, it was confirmed that there were no missing values, so no further treatment was necessary. Upon completion, the cleaned dataset was reduced to three variables: Time, Season, and Total Deaths, as few samples are outlined in @tbl-sample.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-sample
#| tbl-cap: "Samples of the cleaned data"

# Display few of samples.
kable(head(analysis_data), col.names = c("Time", "Season", "Number of decedents"))
```

## Statistical summary of the cleaned data
@tbl-sumarry provides some basic statistics about the dataset. The project spans a total of 204 months. During this period, the highest number of deaths recorded in a single month was 19, while the lowest was 0. The average number of deaths per month is 3.848. Additionally, we can infer that the data is somewhat skewed, as the highest number of deaths in a single month is 19, while the average and median number of deaths are 3 and 3.83, respectively.
```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-sumarry
#| tbl-cap: "Summary of the cleaned data"

# Showing the summary of cleaned data.
kable(summary(analysis_data[, c("year_month", "total_decedents")]),
      col.names = c("Time", "Number of decedents"))
```

@fig-decedentdis illustrates the distribution of monthly shelter resident deaths, showing a left-skewed normal distribution. This indicates that the majority of months have a death count in the range of 0-5. Additionally, there are two outliers, where the death count exceeds 15 in two months, which is an uncommon occurrence.

```{r, fig.pos="H"}
#| warning: false
#| message: false
#| echo: false
#| label: fig-decedentdis
#| fig-cap: "Distribution of monthly number of decedents"

# Plotting a histogram for total_decedents
ggplot(analysis_data, aes(x = total_decedents)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of monthly number of decedents", x = "Number of decedents", y = "Frequency") +
  theme_minimal()+
  theme(
    panel.grid.major = element_line(color = "gray80"),  
    panel.grid.minor = element_blank(),  
    plot.title = element_text(hjust = 0.5, size = 12),  
    axis.title.x = element_text(size = 12), 
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(hjust = 0.5,  size = 10),
    axis.text.y = element_text(hjust = 0,  size = 10),
  )
```

@fig-deathovertime depicts the monthly variation in the number of deaths over time. The overall trend reveals a significant increase in the total number of shelter resident deaths in Toronto, particularly after 2018. Prior to 2018, the number of deaths remained relatively stable, rarely exceeding 5 per month. However, from 2018 onwards, there is not only a noticeable upward trend in deaths but also increased volatility. Spikes in the data become more frequent, with several months recording over 10 deaths, and some months reaching as high as 15 or more.

```{r, fig.pos="H"}
#| warning: false
#| message: false
#| echo: false
#| label: fig-deathovertime
#| fig-cap: "Number of decedents over time"

# plot an graph for the trend of number of decedents over time.
ggplot(analysis_data, aes(x = as.Date(paste(year_month, "-01", sep="")), y = total_decedents)) +
  geom_line(color = "steelblue", size = 0.7) +  
  labs(title = "Number of decedents over time", 
       x = "Time", 
       y = "Number of decedents") +
  scale_x_date(date_labels = "%Y", date_breaks = "2 years") +  
  theme_minimal(base_size = 15) +  
  theme(
    panel.grid.major = element_line(color = "gray80"),  
    panel.grid.minor = element_blank(),  
    plot.title = element_text(hjust = 0.5, size = 12),  
    axis.title.x = element_text(size = 12), 
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 30, hjust = 0.5,  size = 10),
    axis.text.y = element_text(hjust = 0,  size = 10),
  )
```
## Discussion of measurement

The "Deaths of Shelter Residents" dataset is collected by Toronto Public Health (TPH), a governmental body (@Opendatatoronto). Information is sourced from several avenues, including a secure online form where agencies supporting homeless and under-housed populations report deaths (@Opendatatoronto). Other sources include the Toronto Shelter & Support Services (TSSS) division and monthly reports from the Toronto Homeless Memorial. Any duplicate entries are removed, and data are cross-checked by the Office of the Chief Coroner of Ontario (OCCO) (@Opendatatoronto).

Given that this dataset originates from major institutions in Toronto, its credibility is very high. The additional verification by the Office of the Chief Coroner of Ontario (OCCO) further enhances the reliability of the data.

Additionally, collecting data on the deaths of shelter residents is more reliable compared to tracking the entire population. First, Toronto is a large metropolitan area, where the presence of witnesses makes data falsification or omission nearly impossible. Second, tracking deaths is more dependable than attempting to count the entire population of shelter residents, as this group tends to be less organized, more transient, and harder to communicate with. As a result, it is not feasible to accurately account for the entire population in detail (@EndingH).

@@

I used the total number of shelter resident deaths per month as the target variable to be predicted, with time and season as predictors, and performed a linear regression analysis. Since time and season are evidently highly correlated, I decided to retain only time as the final predictor for the model. After completing the simple linear regression, I obtained a positive slope, represented by a straight line (@fig-slr), indicating that the number of shelter resident deaths in Toronto has increased over time. A comparison with the actual data revealed that the real values closely align with the predictions from the SLR model, demonstrating the reliability of my model.

```{r, results='hide', fig.pos="H"}
#| warning: false
#| message: false
#| echo: false
#| label: fig-slr
#| fig-cap: "Simple linear regression of number of decedent by time"

# Scatter plot and fitted line
ggplot(analysis_data, aes(x = ym(year_month), y = total_decedents)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Simple linear regression of number of decedent by time",
       x = "Time",
       y = "Number of Decedents") +
  theme(
    panel.grid.major = element_line(color = "gray80"),  
    panel.grid.minor = element_blank(),  
    plot.title = element_text(hjust = 0.5, size = 12),  
    axis.title.x = element_text(size = 12), 
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(hjust = 0.5,  size = 10),
    axis.text.y = element_text(hjust = 0,  size = 10),
  )
```

@@

```{r, fig.pos="H"}
#| warning: false
#| message: false
#| echo: false
#| label: fig-yeardeath
#| fig-cap: "Number of decedents per Year"

@@

# Create a bar chart with percentage labels
ggplot(filtered_data, aes(x = year, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_text(aes(label = paste0(round(total_decedents_pct, 1), "%")), 
            vjust = -0.5, size = 2.3) +  # Add percentage labels
  labs(title = "Number of decedents per year", x = "Year", y = "Number of decedents") +
  theme_minimal()+
  theme(
    panel.grid.major = element_line(color = "gray80"),  
    panel.grid.minor = element_blank(),  
    plot.title = element_text(hjust = 0.5, size = 12),  
    axis.title.x = element_text(size = 12), 
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 30, hjust = 0.5,  size = 10),
    axis.text.y = element_text(hjust = 0,  size = 10),

)
```

@@

```{r, fig.pos="H"}
#| warning: false
#| message: false
#| echo: false
#| label: fig-seasondeath
#| fig-cap: "Average number of decedents per season"

@@

# Rearranging season.
seasonal_data$season <- factor(seasonal_data$season, levels = c("Spring", "Summer", "Autumn", "Winter"))
seasonal_data_sorted <- seasonal_data %>% arrange(season)

# Create a bar chart with percentage labels
ggplot(seasonal_data, aes(x = season, y = mean_decedents, fill = season)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(total_decedents_pct, 1), "%")), 
            vjust = -0.5, size = 2.5) +  # Add percentage labels
  labs(title = "Average number of decedents per season", 
       x = "Season", 
       y = "Average number of decedents") +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(color = "gray80"),  
    panel.grid.minor = element_blank(),  
    plot.title = element_text(hjust = 0.5, size = 12),  
    axis.title.x = element_text(size = 12), 
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(hjust = 0.5,  size = 10),
    axis.text.y = element_text(hjust = 0,  size = 10),
  )
```
